
# This is part 1 of Project 2

# This file contains code to create the following plots: Heatmap with xCELL results for all samples + cell subtypes based on CD biomarkers expression + overall survival data; Kaplan-Meier plots for B cell subtypes survival analysis (high and low abundances); Cox proportional Hazards Model for multivariative analysis of B cell subtypes and overall survival; Group Comparisons for Class-switched and pro B cells across all tumor sites and molecular clusters;



```{r}
# Step 1: import libraries for heatmap


library(ComplexHeatmap)
library(dplyr)
library(tidyverse)
library(survival)
library(survminer)
library(colorRamp2)
```

### Heatmap of cell-types in SU2C - tumour biopsies

```{r}
# Step 2: Import B cells data for heatmap and modify


corr.df <- read_tsv('/Users/d/Desktop/Life/UK/Imperial/MRes/Assignments/Project_2/BCR_SU2C/XCELL_correlation_analysis/b_cells_surv_xcell.df.txt')

pro_b_cell_data <- read_tsv('/Users/d/Desktop/Life/UK/Imperial/MRes/Assignments/Project_2/SU2C_BCR_clinical_relevance/SU2C.cell.prop.df.txt')

pro_b_cell_data <- as.data.frame(pro_b_cell_data)

rownames(pro_b_cell_data) <- pro_b_cell_data$cell_type
pro_b_cell_data$cell_type <- NULL

```

```{r}

# Step 3: additional data modifications and row selections

common_columns <- intersect(names(RNA.seq.df), names(pro_b_cell_data))
RNA.seq.df_common <- RNA.seq.df[, common_columns, drop = FALSE]
pro_b_cell_data_common <- pro_b_cell_data[, common_columns, drop = FALSE]

df1_non_common_columns <- setdiff(names(RNA.seq.df), common_columns)
df2_non_common_columns <- setdiff(names(pro_b_cell_data), common_columns)

corr.df_modified <- corr.df[!rownames(corr.df) %in% df2_non_common_columns, , drop = FALSE]

```




```{r celltype-heatmap,results="hide", fig.align='center', fig.height=6, fig.width=10,fig.cap="This heatmap represents cell-types in SU2C samples at the tumour site." }

# Step 4: create a final dataframe with a specific columns 

cell.type.cols = list('CD19'=colorRamp2(c(-1:5), hcl_palette = "Reds",reverse = T),
                      'CD24'=colorRamp2(c(-1:5), hcl_palette = "Reds",reverse = T),
                      'CD27'=colorRamp2(c(-1:5), hcl_palette = "Reds",reverse = T),
                      'CD38'=colorRamp2(c(-1:5), hcl_palette = "Reds",reverse = T),
                      'OS'=colorRamp2(c(0:10), hcl_palette = "heat",reverse = T))

tmp.df <- na.omit(corr.df_modified)
colAnnot.df = columnAnnotation(
    'OS'=(corr.df_modified$OS),
    'CD19'=scale(corr.df_modified$CD19),
    'CD27'=scale(corr.df_modified$CD27),
    'CD38'=scale(corr.df_modified$CD38),
    annotation_height=unit(c(rep(5,3)), "cm"),
    col=cell.type.cols,show_legend=T,border=F,na_col='white',annotation_legend_param = list(direction='horizontal'),
    annotation_name_gp= gpar(fontsize = 10,fontface='bold'))
ht_opt$COLUMN_ANNO_PADDING = unit(0.1, "lines")
rownames(corr.df_modified) <- corr.df_modified$Sample.ID

RNA.seq.df <- as.data.frame(t(corr.df_modified[-c(1:4,10:11)]))
colnames(RNA.seq.df) <- corr.df_modified$Sample.ID
# col.ord <- names(sort(colSums(t(scale(t(RNA.seq.df)))[c("CD4 Tcell","CD8 Tcell","Tregs","T cell (other)"),]),decreasing = T))
col.ord <- arrange(tmp.df,CD19)$Sample.ID
```


```{r}

# Step 5: sligthly clean the data to prepare for proper visualization and naming

full_data_abundance <- rbind(RNA.seq.df, pro_b_cell_data_common)
full_data_abundance <- as.matrix(full_data_abundance)
selected_rownames <- c("B cell", "B cell naive", "B cell plasma", "Class-switched memory B cell","B cell memory","pro B-cells")

full_data_abundance_final <- full_data_abundance[rownames(full_data_abundance) %in% selected_rownames, , drop = FALSE]

old_name <- "Class-switched memory B cell"
rownames(full_data_abundance_final)[rownames(full_data_abundance_final) == old_name] <- "Class-switched B cell"

```


```{r celltype-heatmap,results="hide", fig.align='center', fig.height=6, fig.width=10,fig.cap="This heatmap represents cell-types in SU2C samples at the tumour site." }

# Step 6: draw a heatmap to visualize all samples CD biomarkers expression, B cells abundancies and OS


HeatMapFull <- draw(Heatmap(matrix = full_data_abundance_final,
             col = colorRamp2(c(0:5), hcl_palette = "Oslo",reverse = F),
             name = 'Normalized cell prop',border = T,heatmap_legend_param = list(direction='vertical'),
             na_col ='white',
             column_order = col.ord,cluster_columns = F,
             row_title = "Cell proportion",
             show_column_dend = T,
             row_dend_reorder = F,cluster_rows = F,row_title_side = 'left',
             row_dend_width = unit(10,'mm'),row_dend_side = 'left',
             rect_gp = gpar(col = "white", lwd = 0),show_heatmap_legend = T,show_parent_dend_line = T,
             show_row_dend = F,show_column_names = F,
             show_row_names = T,cluster_row_slices = T,
             width = nrow((full_data_abundance_final))*unit(1.7, "cm"),
             height = ncol((full_data_abundance_final))*unit(0.2, "mm"),column_title = 'Tumour samples',column_title_side = 'bottom',
             row_names_gp = gpar(fontsize=10,fontface='bold'),cluster_column_slices = F,column_dend_reorder = T,column_title_gp = gpar(fontsize=8,fontface='bold'),
             clustering_distance_rows ="pearson",row_title_gp = gpar(fontsize=8,fontface='bold'),
             top_annotation  = colAnnot.df
),heatmap_legend_side='left',annotation_legend_side='bottom')

```


```{r}
# Step 7: Import data for Kaplan-Meier survival analysis


surv_b_cells <- read_csv('/Users/d/Desktop/Life/UK/Imperial/MRes/Assignments/Project_2/BCR_SU2C/B_cells_subtypes_survival.csv')
```

```{r}

surv_b_cells <- surv_b_cells %>% dplyr::rename(B_cells = `B-cells`, Class_switched_memory_B_cells = `Class-switched memory B-cells`, Memory_B_cells = `Memory B-cells`, naive_B_cells= `naive B-cells`, pro_B_cells = `pro B-cells` )

surv_b_cells$OS_days <- surv_b_cells$OS * 365

```



```{r}
# Step 8: create Kaplan-Meier Curves for each type of B cells separately
# Part 1: B_cells

surv_object <- Surv(time = surv_b_cells$OS_days, event = surv_b_cells$death)

fit <- survfit(surv_object ~ B_cells, data = surv_b_cells)

kaplan_b_cells <- ggsurvplot(fit, data = surv_b_cells, 
             pval = TRUE, 
             conf.int = TRUE, 
             risk.table = TRUE,
             ggtheme = theme_minimal(),
             xlab = "",
             ylab = "")

summary(survfit(Surv(time = surv_b_cells$OS_days, event = surv_b_cells$death) ~ surv_b_cells$`B_cells`), times = round(c(365 / 2), digits = 0))
```


```{r}

# Step 8: create Kaplan-Meier Curves for each type of B cells separately
# Part 2: Class_switched_memory_B_cells

surv_object <- Surv(time = surv_b_cells$OS_days, event = surv_b_cells$death)

fit <- survfit(surv_object ~ Class_switched_memory_B_cells, data = surv_b_cells)

kaplan_class_b_cells <- ggsurvplot(fit, data = surv_b_cells, 
             pval = TRUE, 
             conf.int = TRUE, 
             risk.table = TRUE,
             ggtheme = theme_minimal())


summary(survfit(Surv(time = surv_b_cells$OS_days, event = surv_b_cells$death) ~ surv_b_cells$`Class_switched_memory_B_cells`), times = round(c(365 / 2), digits = 0))
```
```{r}
# Step 8: create Kaplan-Meier Curves for each type of B cells separately
# Part 3: Memory_B_cells

surv_object <- Surv(time = surv_b_cells$OS_days, event = surv_b_cells$death)

fit <- survfit(surv_object ~ Memory_B_cells, data = surv_b_cells)

kaplan_memory_b_cells <- ggsurvplot(fit, data = surv_b_cells, 
             pval = TRUE, 
             conf.int = TRUE, 
             risk.table = TRUE,
             ggtheme = theme_minimal())

summary(survfit(Surv(time = surv_b_cells$OS_days, event = surv_b_cells$death) ~ surv_b_cells$`Memory_B_cells`), times = round(c(365 / 2), digits = 0))
```


```{r}

# Step 8: create Kaplan-Meier Curves for each type of B cells separately
# Part 4: naive_B_cells

surv_object <- Surv(time = surv_b_cells$OS_days, event = surv_b_cells$death)

fit <- survfit(surv_object ~ naive_B_cells, data = surv_b_cells)

kaplan_naive_b_cells <- ggsurvplot(fit, data = surv_b_cells, 
             pval = TRUE, 
             conf.int = TRUE, 
             risk.table = TRUE,
             ggtheme = theme_minimal())

summary(survfit(Surv(time = surv_b_cells$OS_days, event = surv_b_cells$death) ~ surv_b_cells$`naive_B_cells`), times = round(c(365 / 2), digits = 0))
```

```{r}

# Step 8: create Kaplan-Meier Curves for each type of B cells separately
# Part 5: pro_B_cells

surv_object <- Surv(time = surv_b_cells$OS_days, event = surv_b_cells$death)

fit <- survfit(surv_object ~ pro_B_cells, data = surv_b_cells)

kaplan_pro_b_cells <- ggsurvplot(fit, data = surv_b_cells, 
             pval = TRUE, 
             conf.int = TRUE, 
             risk.table = TRUE,
             ggtheme = theme_minimal())

summary(survfit(Surv(time = surv_b_cells$OS_days, event = surv_b_cells$death) ~ surv_b_cells$`pro_B_cells`), times = round(c(365 / 2), digits = 0))

```


```{r}

# Step 9: Cox Proportional Hazard Model - multivariative analysis on B cell subtypes abundance effects on OS

surv_b_cells_no_id <- surv_b_cells[ , -which(names(surv_b_cells) == "cell_type")]
surv_object <- Surv(time = surv_b_cells_no_id$OS, event = surv_b_cells_no_id$death)
cox_model <- coxph(surv_object ~ B_cells + naive_B_cells + Class_switched_memory_B_cells + pro_B_cells + Memory_B_cells, data = surv_b_cells_no_id)
summary(cox_model)

forest_plot <- ggforest(cox_model, data = surv_b_cells_no_id,
         main = "Cox Proportional Hazards Model (multivariative) for B cell subtypes in SU2C cohort",
         cpositions = c(0.01, 0.22, 5.4),
         fontsize = 0.8,
         refLabel = 'reference')

ggsave("forest_plot.png", plot = forest_plot, width = 12, height = 6)


```

```{r}

# Step 10: load, prepare, merge and transform data for B cell abundance group comparisons across localizations and clusters

cd_b_cells_data_for_distribution <- cd_b_cells_data
cd_b_cells_data_for_distribution$Sample.ID <- paste0(cd_b_cells_data_for_distribution$Sample.ID, "_report")

xcell_abundance_distribution_local <- merge(cd_b_cells_data_for_distribution, clin_data_important, by.x = "Sample.ID", by.y = "POLY_A", all.x = TRUE )

xcell_abundance_distribution_full <- merge(xcell_abundance_distribution_local, clusters_df, by.x = "Sample.ID", by.y = "Sample_ID", all.x = TRUE )

xcell_abundance_distribution_full <- xcell_abundance_distribution_full %>% dplyr::rename(B_cells = `B cell`, Class_switched_memory_B_cells = `Class-switched memory B cell`, Memory_B_cells = `B cell memory`, naive_B_cells= `B cell naive`, pro_B_cells = `B cell plasma` )

xcell_abundance_distribution_full$clusters <- factor(xcell_abundance_distribution_full$clusters, levels = c("1", "2", "3", "4"))

xcell_abundance_distribution_full <- na.omit(xcell_abundance_distribution_full)


min_value_cs_b <- min(xcell_abundance_distribution_full$Class_switched_memory_B_cells)
xcell_abundance_distribution_full$Class_switched_memory_B_cells_shifted <- xcell_abundance_distribution_full$Class_switched_memory_B_cells + abs(min_value_cs_b) + 1e-10 


min_value_pro_b <- min(xcell_abundance_distribution_full$pro_B_cells)
xcell_abundance_distribution_full$pro_B_cells_shifted <- xcell_abundance_distribution_full$pro_B_cells + abs(min_value_pro_b) + 1e-10 


custom_colors_localization <- c("Lymph_node" = "#1f77b4",  # Blue
                   "Bone" = "#ff7f0e",        # Orange
                   "Liver" = "#2ca02c",       # Green
                   "Prostate" = "#d62728",    # Red
                   "Other" = "#9467bd")       # Purple

custom_colors_clusters <- c("#1f77b4", "#ff7f0e", 'darkgreen','purple')

```

```{r}

# Step 11: Tumor site and cluster comparison with statistics - Class-switched B cells

xcell_localization_cs_b_cell <- ggplot(xcell_abundance_distribution_full, aes(x = Biopsy_Site, y = Class_switched_memory_B_cells_shifted, fill = Biopsy_Site)) +
  geom_boxplot() +
  theme_minimal() +
  stat_compare_means(method = "wilcox.test", 
                     comparisons = list(c("Lymph_node", "Bone"), c("Lymph_node", "Liver"), c("Lymph_node", "Prostate"), c("Lymph_node", "Other")),
                     label = "p.signif") +
  stat_compare_means(method = "kruskal.test", label = "p.format", vjust = -0.3, hjust = -4) +
  labs(title = "Class-switched B cells abundance across metastasis localization",
       x = "Localization",
       y = "Relative Abundance",
       fill = "Metastasis Localization") +
  theme(
  plot.title = element_text(face = "bold"),
  axis.title.x = element_text(face = "bold"),
  axis.title.y = element_text(face = "bold"),
  legend.title = element_text(face = "bold"),
  axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = custom_colors_localization)


wilcox_results_cs_b_cells_localization <- compare_means(Class_switched_memory_B_cells_shifted ~ Biopsy_Site, data = xcell_abundance_distribution_full,
                                method = "wilcox.test",
                               comparisons = list(c("Lymph_node", "Bone"), c("Lymph_node", "Liver"), c("Lymph_node", "Prostate"), c("Lymph_node", "Other")))


kruskal_results_cs_b_cells_localization <- compare_means(Class_switched_memory_B_cells_shifted ~ Biopsy_Site, data = xcell_abundance_distribution_full, method = "kruskal.test")


descriptive_stats_cs_b_cells_localization <- xcell_abundance_distribution_full %>%
  group_by(Biopsy_Site) %>%
  summarise(
    count = n(),
    mean = mean(Class_switched_memory_B_cells_shifted, na.rm = TRUE),
    median = median(Class_switched_memory_B_cells_shifted, na.rm = TRUE),
    sd = sd(Class_switched_memory_B_cells_shifted, na.rm = TRUE),
    min = min(Class_switched_memory_B_cells_shifted, na.rm = TRUE),
    max = max(Class_switched_memory_B_cells_shifted, na.rm = TRUE),
    IQR = IQR(Class_switched_memory_B_cells_shifted, na.rm = TRUE)
  )



xcell_clusters_cs_b_cell <- ggplot(xcell_abundance_distribution_full, aes(x = clusters, y = Class_switched_memory_B_cells_shifted, fill = clusters)) +
  geom_boxplot() +
  theme_minimal() +
  stat_compare_means(method = "wilcox.test", 
                     comparisons = list(c("1", "2"), c("1", "3"), c("1", "4")),
                     label = "p.signif") +
  stat_compare_means(method = "kruskal.test", label = "p.format", vjust = -0.3, hjust = -4) +
  labs(title = "Class-switched B cells abundance across molecular clusters",
       x = "Clusters",
       y = "Relative Abundance",
       fill = "Clusters") +
  theme(
  plot.title = element_text(face = "bold"),
  axis.title.x = element_text(face = "bold"),
  axis.title.y = element_text(face = "bold"),
  legend.title = element_text(face = "bold"),
  axis.text.x = element_text(angle = 45, hjust = 1)) +
scale_fill_manual(values = custom_colors_clusters)

wilcox_results_cs_b_cells_clusters <- compare_means(Class_switched_memory_B_cells_shifted ~ clusters, data = xcell_abundance_distribution_full,
                                method = "wilcox.test",
                                comparisons = list(c("1", "2"), c("1", "3"), c("1", "4")))


kruskal_results_cs_b_cells_clusters <- compare_means(Class_switched_memory_B_cells_shifted ~ clusters, data = xcell_abundance_distribution_full, method = "kruskal.test")


descriptive_stats_cs_b_cells_clusters <- xcell_abundance_distribution_full %>%
  group_by(clusters) %>%
  summarise(
    count = n(),
    mean = mean(Class_switched_memory_B_cells_shifted, na.rm = TRUE),
    median = median(Class_switched_memory_B_cells_shifted, na.rm = TRUE),
    sd = sd(Class_switched_memory_B_cells_shifted, na.rm = TRUE),
    min = min(Class_switched_memory_B_cells_shifted, na.rm = TRUE),
    max = max(Class_switched_memory_B_cells_shifted, na.rm = TRUE),
    IQR = IQR(Class_switched_memory_B_cells_shifted, na.rm = TRUE)
  )

```

```{r}

# Step 12: Tumor site and cluster comparison with statistics - pro B cells


xcell_localization_pro_b_cell <- ggplot(xcell_abundance_distribution_full, aes(x = Biopsy_Site, y = pro_B_cells_shifted, fill = Biopsy_Site)) +
  geom_boxplot() +
  theme_minimal() +
  stat_compare_means(method = "wilcox.test", 
                     comparisons = list(c("Lymph_node", "Bone"), c("Lymph_node", "Liver"), c("Lymph_node", "Prostate"), c("Lymph_node", "Other")),
                     label = "p.signif") +
  stat_compare_means(method = "kruskal.test", label = "p.format", vjust = -0.3, hjust = -4) +
  labs(title = "Pro B cells abundance across metastasis localization",
       x = "Localization",
       y = "Relative Abundance",
       fill = "Metastasis Localization") +
  theme(
  plot.title = element_text(face = "bold"),
  axis.title.x = element_text(face = "bold"),
  axis.title.y = element_text(face = "bold"),
  legend.title = element_text(face = "bold"),
  axis.text.x = element_text(angle = 45, hjust = 1)) + 
  scale_fill_manual(values = custom_colors_localization)




wilcox_results_pro_b_cells_localization <- compare_means(pro_B_cells_shifted ~ Biopsy_Site, data = xcell_abundance_distribution_full,
                                method = "wilcox.test",
                               comparisons = list(c("Lymph_node", "Bone"), c("Lymph_node", "Liver"), c("Lymph_node", "Prostate"), c("Lymph_node", "Other")))


kruskal_results_pro_b_cells_localization <- compare_means(pro_B_cells_shifted ~ Biopsy_Site, data = xcell_abundance_distribution_full, method = "kruskal.test")


descriptive_stats_pro_b_cells_localization <- xcell_abundance_distribution_full %>%
  group_by(Biopsy_Site) %>%
  summarise(
    count = n(),
    mean = mean(pro_B_cells_shifted, na.rm = TRUE),
    median = median(pro_B_cells_shifted, na.rm = TRUE),
    sd = sd(pro_B_cells_shifted, na.rm = TRUE),
    min = min(pro_B_cells_shifted, na.rm = TRUE),
    max = max(pro_B_cells_shifted, na.rm = TRUE),
    IQR = IQR(pro_B_cells_shifted, na.rm = TRUE)
  )



xcell_clusters_pro_b_cell <- ggplot(xcell_abundance_distribution_full, aes(x = clusters, y = pro_B_cells_shifted, fill = clusters)) +
  geom_boxplot() +
  theme_minimal() +
  stat_compare_means(method = "wilcox.test", 
                     comparisons = list(c("1", "2"), c("1", "3"), c("1", "4")),
                     label = "p.signif") +
  stat_compare_means(method = "kruskal.test", label = "p.format", vjust = -0.3, hjust = -4) +
  labs(title = "Pro B cells abundance across molecular clusters",
       x = "Clusters",
       y = "Relative Abundance",
       fill = "Clusters") +
  theme(
  plot.title = element_text(face = "bold"),
  axis.title.x = element_text(face = "bold"),
  axis.title.y = element_text(face = "bold"),
  legend.title = element_text(face = "bold"),
  axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_fill_manual(values = custom_colors_clusters)


wilcox_results_pro_b_cells_clusters <- compare_means(pro_B_cells_shifted ~ clusters, data = xcell_abundance_distribution_full,
                                method = "wilcox.test",
                                comparisons = list(c("1", "2"), c("1", "3"), c("1", "4")))

kruskal_results_pro_b_cells_clusters <- compare_means(pro_B_cells_shifted ~ clusters, data = xcell_abundance_distribution_full, method = "kruskal.test")


descriptive_stats_pro_b_cells_clusters <- xcell_abundance_distribution_full %>%
  group_by(clusters) %>%
  summarise(
    count = n(),
    mean = mean(pro_B_cells_shifted, na.rm = TRUE),
    median = median(pro_B_cells_shifted, na.rm = TRUE),
    sd = sd(pro_B_cells_shifted, na.rm = TRUE),
    min = min(pro_B_cells_shifted, na.rm = TRUE),
    max = max(pro_B_cells_shifted, na.rm = TRUE),
    IQR = IQR(pro_B_cells_shifted, na.rm = TRUE)
  )


```

